---
title: "Introduction to Data"
author: "Shravan Kuchkula"
output:
  github_document:
    toc: yes
  html_document:
    keep_md: yes
    theme: cosmo
    toc: yes
---

## Language of data

We will first start by exploring the `high school and beyond` **hsb2** dataset which part of the `openintro` package. Notice here that if we are loading a dataset which is part of a package, we will be using the `data()` function to load the dataset. Other ways to import data into R would be from flatfiles, websites, APIs etc.

As usual, we will use our function to install/load the required packages into this session.

```{r message = FALSE, warning = FALSE}
# Install a package and load it.
installRequiredPackages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[,"Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

# List of packages used in this session
libs <- c("dplyr", "ggplot2", "tidyr", "openintro")

# call the function
installRequiredPackages(libs)
```

Load the dataset `hsb2` 

```{r}
data("hsb2")
```

Glimpse the dataset `hsb2`

```{r}
glimpse(hsb2)
```

Load the UCBAdmissions dataset and store it in ucb_admit.
```{r}
ucb_admit <- data("UCBAdmissions")

str(ucb_admit)

# Count number of male and female applicants admitted
#ucb_counts <- ucb_admit %>%
#  count(gender, Admit)
```
